---
title: "309final"
author: "Mara Augustynowicz"
date: "2026-05-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Part 1 
```{r}
#load data 
beaufort_scale <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-04-14/beaufort_scale.csv')
birds <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-04-14/birds.csv')
sea_states <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-04-14/sea_states.csv')
ships <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-04-14/ships.csv')

#clean
library(dplyr)
birddata <- birds %>%
  inner_join(ships, by = "record_id")

birddata <- birddata %>%
  select(feeding, speed, air_temperature, wind_speed_class, sea_state_class, precipitation)

# remove missing values
birddata <- na.omit(birddata)

# recode variables
birddata <- birddata %>%
  mutate(
    feeding          = factor(ifelse(feeding == TRUE, "Yes", "No")),
    wind_speed_class = factor(wind_speed_class),
    sea_state_class  = factor(sea_state_class),
    precipitation    = factor(precipitation)
  )

# response variable is if feeding or not 
```

### Part 2 
```{r}
library(tidyverse)
library(ggplot2)
library(patchwork)

p1 <- ggplot(birddata, aes(x = speed, y = feeding)) +
  geom_jitter(height = 0.1, alpha = 0.3) +
  labs(x = "Ship Speed",
       y = "Feeding",
       title = "Bird Feeding vs Ship Speed")

p2 <- ggplot(birddata, aes(x = air_temperature, y = feeding)) +
  geom_jitter(height = 0.1, alpha = 0.3) +
  labs(x = "Air Temperature",
       y = "Feeding",
       title = "Bird Feeding vs Air Temperature")

p3 <- ggplot(birddata, aes(x = wind_speed_class, fill = feeding)) +
  geom_bar(position = "fill") +
  labs(x = "Wind Speed Class (Beaufort Scale)",
       y = "Proportion",
       title = "Bird Feeding by Wind Speed Class")

p4 <- ggplot(birddata, aes(x = sea_state_class, fill = feeding)) +
  geom_bar(position = "fill") +
  labs(x = "Sea State Class",
       y = "Proportion",
       title = "Bird Feeding by Sea State")


p5 <- ggplot(birddata, aes(x = precipitation, fill = feeding)) +
  geom_bar(position = "fill") +
  labs(x = "Precipitation Type",
       y = "Proportion",
       title = "Bird Feeding by Precipitation Type")

library(patchwork)
bird_dashboard <-
((p1 | p2) /
(p3 | p4)) /
  p5 +
plot_annotation(
  title = "Bird Feeding",
  subtitle = "Comparing bird feeding and enviromental conditions "
)
bird_dashboard
```

### Part 3
```{r}
library(tidyverse)
library(caret)
library(rpart)
library(randomForest)

set.seed(511)

# split data
birdIndex  <- createDataPartition(birddata$feeding, p = 0.7, list = FALSE)
bird_train <- birddata[ birdIndex, ]
bird_test  <- birddata[-birdIndex, ]

# 5-fold cross-validation setup
cvmethod <- trainControl(
  method          = "cv",
  number          = 5,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary
)

# set positive class
bird_train$feeding <- relevel(bird_train$feeding, ref = "Yes")
bird_test$feeding  <- relevel(bird_test$feeding,  ref = "Yes")

# Logistic Regression (all predictors) 
# Rationale: a baseline that uses every available predictor
fit_glm_all <- train(
  feeding ~ .,
  data      = bird_train,
  method    = "glm",
  family    = "binomial",
  trControl = cvmethod,
  metric    = "ROC"
)

# Logistic Regression (subset of predictors) 
# Rationale: compare to full model, might prevent overfitting
fit_glm_small <- train(
  feeding ~ speed + air_temperature,
  data      = bird_train,
  method    = "glm",
  family    = "binomial",
  trControl = cvmethod,
  metric    = "ROC"
)

# Decision Tree 
# Rationale: is easy to visualize and interpret
fit_tree <- train(
  feeding ~ .,
  data      = bird_train,
  method    = "rpart",
  trControl = cvmethod,
  metric    = "ROC"
)

# Random Forest 
# Rationale: reduces variance
fit_rf <- train(
  feeding ~ .,
  data      = bird_train,
  method    = "rf",
  trControl = cvmethod,
  metric    = "ROC"
)

# k-Nearest Neighbors 
# Rationale: makes no distributional assumptions
fit_knn <- train(
  feeding ~ .,
  data       = bird_train,
  method     = "knn",
  trControl  = cvmethod,
  metric     = "ROC",
  preProcess = c("center", "scale")
)

# view cross-validation results
fit_glm_all$results
fit_glm_small$results
fit_tree$results
fit_rf$results
fit_knn$results
fit_knn$results
```

### Part 4
```{r}
library(caret)
library(pROC)
library(dplyr)
library(tidyr)
library(ggplot2)

# store models in list 
models <- list(
  "Logistic (all)" = fit_glm_all,
  "Logistic (subset)" = fit_glm_small,
  "Decision Tree" = fit_tree,
  "Random Forest" = fit_rf,
  "k-NN" = fit_knn
)

# cross val results
cv_results <- lapply(names(models), function(name) {
  fit <- models[[name]]
  best <- merge(fit$bestTune, fit$results)
  data.frame(
    Model = name,
    ROC = best$ROC[1],
    Sens = best$Sens[1],
    Spec = best$Spec[1]
  )
}) %>%
  bind_rows()


# test results 
test_results <- lapply(names(models), function(name) {
  fit <- models[[name]]
  preds <- predict(fit, newdata = bird_test)
  probs <- predict(fit, newdata = bird_test, type = "prob")[,"Yes"]

  cm <- confusionMatrix(
    preds,
    bird_test$feeding,
    positive = "Yes"
  )

  auc_val <- auc(
    roc(bird_test$feeding, probs,
        levels = c("No", "Yes"))
  )

  data.frame(
    Model = name,
    Accuracy = cm$overall["Accuracy"],
    AUC = as.numeric(auc_val),
    Sensitivity = cm$byClass["Sensitivity"],
    Specificity = cm$byClass["Specificity"]
  )

}) %>%
  bind_rows()


# visualization
test_long <- test_results %>%
  pivot_longer(
    cols = c(Accuracy, AUC),
    names_to = "Metric",
    values_to = "Value"
  )

ggplot(test_long,
       aes(x = Model,
           y = Value,
           fill = Metric)) +
  geom_col(position = "dodge") +
  ylim(0,1) +
  labs(
    title = "Model Performance Comparison",
    y = "Performance Score"
  ) +
  theme_minimal()
```

The random forest performed the best because it had the highest AUC and accuracy, but there is still a risk of overfitting. The predictions from these models could help researchers better understand the environmental conditions associated with bird feeding behavior. Identifying patterns in feeding activity may improve ecological monitoring and support future studies on bird behavior and habitat conditions. 